---
title: "Data_Analysis_DO_per"
author: "Andrew Bazuro"
format: html
editor: visual
---
To start I will first read in the data from the exploration that I plan to use for creating my model

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
Stream = read.csv(here("Data", "Bazuro_Capstone_Final.csv"))
```

After reading in the dataset we should check it to ensure that all variables are the correct types. Since I had to change several in the exploration I would expect to have to change several now.

```{r}
glimpse(Stream)
```
Here we can see that the dataset is made up of 327
My analysis will only use `Concentration_ppm` and `DO_per` so I won't change any variable types.

```{r}
glimpse(Stream)
```

The question that I am examining is: If there is a significant relationship between DO_per and the Concentration_ppm in water.

#Plot
I will use a boxplot here to examine the data
```{r}
ggplot(data = Stream, aes(y = DO_per)) +
  geom_boxplot(outlier.color = "red")
```
I can see several outliers here which appear to be concerning.

I will also make a histogram to check the data 
```{r}
ggplot(data = Stream, aes(y = DO_per)) +
  geom_histogram()
```

Based on these plots I will be removing all values below 85 to eliminate the outliers.

```{r}
Stream <- Stream |> filter(DO_per >= 85)
Stream |> arrange(DO_per)
```
Removing the outliers here also removes all the NA values from the dataset, reducing it to only 335 rows.

After removing the outliers I have remade the plot to show the new data distribution.
```{r}
ggplot(data = Stream, aes(x = DO_per))+
  geom_histogram()
```
I see no issues here although there is a value below 90 which may now be identified as an outlier.

```{r}
ggplot(data = Stream, aes(y = DO_per)) +
  geom_boxplot(outlier.color = "red")
```
There is still one outlier but it only became an outlier after I removed the original ones so I will leave this one for now. 

As these are both continous variables I will create a scatterplot to examine the mean and intercept.
```{r}
ggplot(Stream, aes(x = DO_per, y = Concentration_ppm)) +
  geom_point() +
  theme_bw()
```
Here the mean appears to be around 50-100 with an intercept near 0 and a slope near or at 0. However, the large spread leads me to expect that there won't be a significant association between the two.
#Model
I will now create the model that I will be using for this analysis.
```{r}
mod1 = lm(Concentration_ppm ~ DO_per, data = Stream)
```
I have chosen to use a regression here as I am examining a ratio response variable along with a continous predictor. Since there is only a single predictor a regression is ideal.

#Model Assumption
I will now run the `autoplot()` function to examine the conditions.
```{r}
autoplot(mod1)
```
There are some outliers but overall I see no issues here.
#Interpretation
I will now run the `anova()` function to interpre the model
```{r}
anova(mod1)
```
Based on this plot I can see that there is a significant association between the DO_per and Concentration_ppm.

I will now run the `summary()` table for closer examination
```{r}
summary(mod1)
```
Here we can see that the slope for this is around -5 and the intercept appears to be at 475 (note the break in the graph above which I missed earlier). As our R^2 adjusted is 0.050608 which means that 0.05% of our variation is associated with DO_per.

#Plot
```{r}
ggplot(Stream, aes(x = DO_per, y = Concentration_ppm)) +
  geom_point() +
  geom_smooth(method = "lm")
  theme_bw()
```
#Results
There appears to be a statistically significant and negative association between DO_per and Concentration_ppm (F = 18.403, p = 2.361E-05, R^2 = 0.05359)