---
title: "Data_Analysis_DO_per"
author: "Andrew Bazuro"
format: html
editor: visual
---

To start I will first read in the data from the exploration that I plan to use for creating my model

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
Stream = read.csv(here("Data", "Bazuro_Capstone_Final.csv"))
```

After reading in the dataset we should check it to ensure that all variables are the correct types. Since I had to change several in the exploration I would expect to have to change several now.

```{r}
glimpse(Stream)
```

Here we can see that the dataset is made up of 327 My analysis will only use `Concentration_ppm` and `DO_per` so I won't change any variable types. However, since I only want to examine the Fe element I will remove the elements that aren't Fe.

```{r}
Stream <- Stream |> filter(Element == "Fe")
glimpse(Stream)
```

The question that I am examining is: If there is a significant relationship between DO_per and the Concentration_ppm of Fe in water.

#Plot I will use a boxplot here to examine the data

```{r}
ggplot(data = Stream, aes(y = DO_per)) +
  geom_boxplot(outlier.color = "red")
```

I can see several outliers but there is no reason to doubt that they are valid data so I will leave them in.

I will also make a histogram to check the data

```{r}
ggplot(data = Stream, aes(y = DO_per)) +
  geom_histogram()
```
Again while there are outliers I will be leaving them in as they are valid data points and could offer imortant conclusions to my data.

As these are both continous variables I will create a scatterplot to examine the mean and intercept.

```{r}
ggplot(Stream, aes(x = DO_per, y = Concentration_ppm)) +
  geom_point() +
  theme_bw()
```

Here the intercept is near 50 with a negative slope of slope of less than one . However, the large spread leads me to expect that there won't be a significant association between the two. 


#Model 

I will now create the model that I will be using for this analysis.

```{r}
mod1 = lm(Concentration_ppm ~ DO_per, data = Stream)
```

I have chosen to use a regression here as I am examining a ratio response variable along with a continous predictor. Since there is only a single predictor a regression is ideal. 

#Model Assumption I will now run the `autoplot()` function to examine the conditions.

```{r}
autoplot(mod1)
```
Due to the normality issues I will be performing this analysis using glm to allow for a more flexible model.

```{r}
mod1 = glm(Concentration_ppm ~ DO_per, data = Stream)
```

#Interpretation 

I will now run the `anova()` function to interpret the model

```{r}
anova(mod1)
```

Based on this plot I can see that there is a significant association between the DO_per and Concentration_ppm.

I will now run the `summary()` table for closer examination

```{r}
summary(mod1)
```

Here we can see that the slope for this is around -8 and the intercept appears to be at 790.
#Plot

```{r}
ggplot(Stream, aes(x = DO_per, y = Concentration_ppm)) +
  geom_point() +
  geom_smooth(method = "glm")
  theme_bw()
```

#Results There appears to be a statistically significant and negative association between DO_per and Concentration_ppm (F = 55.207, p = 5.68E-11)
