---
title: "Data_Analysis_Element_Type"
author: "Andrew Bazuro"
format: html
editor: visual
---

To start I will first read in the data from the exploration that I plan to use for creating my model

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
Stream = read.csv(here("Data", "Bazuro_Capstone_Final.csv"))
```

After reading in the dataset we should check it to ensure that all variables are the correct types. Since I had to change several in the exploration I would expect to have to change several now.

```{r}
glimpse(Stream)
```

My analysis will use the `Type` and `Element` variables so I will change those to factors for later analysis. There are currently 17 columns in the dataframe and 832 rows left so the remaining data should only include the Elements that we care about.

```{r}
Stream$Element <- as.factor(Stream$Element) #changes the Element variable into a factor
Stream$Type <- as.factor(Stream$Type) #changes the Type variable into a factor
glimpse(Stream)
```

I will also check the levels for both element and type to make sure that I only have the ones I want for the analysis

```{r}
levels(Stream$Element)
levels(Stream$Type)
```

We have four levels in the `Element` column and six in the `Type` column so we are good to go. The question that I am examining is: What is the effect of different habitat types on the abundance of Mg, Na, Fe, and K.

# Plots

## Mg plot We already examined the plots for outliers in the Data exploration so I won't do that here. However, I will still place a histogram here to estimate the means of each section.

```{r}
Stream |> filter(Element == "Mg") |> ggplot(aes(Concentration_ppm))+
  geom_histogram()+
  facet_wrap(~Type)
```

Here I see no outliers and will now create a boxplot to examine the mean.

```{r}
Stream |> filter(Element == "Mg") |> ggplot(aes(Concentration_ppm))+
  geom_boxplot()+
  facet_wrap(~Type)
```

The mean for most graphs is close to zero, while the median in the seep is closer to 200. As such, I expect to see a significant increase in the seep compared to the other habitat types.

## Na plot I will now examine `Na`

```{r}
Stream |> filter(Element == "Na") |> ggplot(aes(Concentration_ppm))+
  geom_histogram()+
  facet_wrap(~Type)
```

Here I do see some outlier but as I have no evidence that they aren't valid data points I will move forward.

```{r}
Stream |> filter(Element == "Na") |> ggplot(aes(Concentration_ppm))+
  geom_boxplot()+
  facet_wrap(~Type)
```

Based on the means I would estimate there to be several significant differences. Notably, seep_control has a lower mean than all the other habitat types. Tributary, river, and lake all appear to have similar means at 1.8 and most likely won't be significantly different. However, wetland does have a higher mean of 2.5 and will most likely be significantly different from the other habitat types. Finally, the seep appears to have a mean of 2.3 and should be significantly different from all the types aside from the wetland.

##K plot I will now examine `K`

```{r}
Stream |> filter(Element == "K") |> ggplot(aes(Concentration_ppm))+
  geom_histogram()+
  facet_wrap(~Type)
```

Here the values are extremely close and while there are some outliers I have already trimmed out the ones that aren't valid data points. As such I will now create a boxplot to examine means.

```{r}
Stream |> filter(Element == "K") |> ggplot(aes(Concentration_ppm))+
  geom_boxplot()+
  facet_wrap(~Type)
```

Based on this plot I would expect a significant difference between the seep and the other plots as it has a mean around 1 while the rest are all closer to 0.25. I would also expect a significant difference between tributary and the other plots as its outliers should significantly change its mean from the other plots.

##Fe plot I will now examine `Fe`

```{r}
Stream |> filter(Element == "Fe") |> ggplot(aes(Concentration_ppm))+
  geom_histogram()+
  facet_wrap(~Type)
```

I see no outliers here so I will now examine the boxplot.

```{r}
Stream |> filter(Element == "Fe") |> ggplot(aes(Concentration_ppm))+
  geom_boxplot()+
  facet_wrap(~Type)
```

While there are again some outliers here they shouldn't be a concern. Based on the means I would expect to see a significant difference in only the seep as it has a mean closer to 150 while the rest are close to 0.

# Model 

As I see no outliers and have examined each plot I will now create the model.


Our null hypothesis is "There is no relationship between `Element`, `Type`, and the `Concentration_ppm`."

Our alternative hypothesis is "There is a relationship between `Element`, `Type`, and the `Concentration_ppm` with


```{r}
mod1 = glm(Concentration_ppm ~ Type * Element, data = Stream) ##concentration_ppm explained by Type * Element. I am using an interaction term to see if the effect of the habitat or element depends on the other element/habitat in question.
```

I have chosen to perform a two-way anova here because I want to examine the effects of two categorical predictor variables on one ratio response variable. Since my samples are both independent an Anova is the best test and will be two tailed as I mainly want to see the difference and can perform more analysis to determine the direction of the difference.

# Model assumptions I will now check the model assumptions using the `autoplot()` function

```{r}
autoplot(mod1) ##When I ran this I noticed a ton of points coming off the line and went back to my exploration to remove way more of the outliers. However, even after doing so I am still seeing the points move off the line. If you think I should go  back and be even stricter please let me know as while I want to keep as many in as possible to account for the different habitat types and Elements this does look concerning.
```

It is noteable here that there are several values near the end of the normality plot that have moved far off the line. This shows that there are still some potential issues with normality but that they aren't very severe and we can move forward.


#Interpretation I will now run the `anova()` function to interpret the model

```{r}
anova(mod1, test = "Chisq")
```

Here we can see that there is a highly significant effect of type and element on the concentration_ppm, as well as a significant interaction between the two main effects. There are 5 df for the type variable because it has 6 levels and we are calcuating n = 6 so df = n - 1 = 5. There are 3 df for element because it has 4 levels and we are calculating n = 4 so df = n - 1 = 3.

I will now run the `summary()` function to examine how the concentration_ppm compares to the intercept for each habitat/type interaction.

```{r}
summary(mod1)
```

Our intercept here is the ElementFe predictor which is significantly different from the seep type and the three other elements.
```{r}
phc1 <- glht(mod1, linfct = mcp(Type = "Tukey"))
phc2 <- glht(mod1, linfct = mcp(Element = "Tukey"))
phc1 <- glht(mod1, linfct = mcp((Type = "Tukey") + (Element = "Tukey")))
summary(phc1)
summary(phc2)
```


I assume that you want us to run a tukeyHSD here but when I attempted to do so I had some errors trying to figure it out. I ended up checking chapter 6 but didn't see anything about how to use a TukeyHSD here so if you'd like one in the final analysis I'd prefer to meet with you about the proper setup for a two-way anova tukey. I ended up following the chapter to create a plot using the means and std deviations to show variation.

I will now create a plot using the means and std deviations of each of the commbinations.

```{r}
sum_stream <- Stream |> group_by(Element, Type) |> summarise(
  mean_Concentration = mean(Concentration_ppm),
    sd_Concentration = sd(Concentration_ppm)/sqrt(n())
)
sum_stream
```

#Plot I will now create the plot

```{r}
ggplot(sum_stream, aes(x = Type, y = mean_Concentration, colour = Element, group = Element)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = mean_Concentration - sd_Concentration,
                    ymax = mean_Concentration + sd_Concentration), width = 0.1) +
  theme_bw()
```

This plot is based off the one from chapter 6. It looks like it runs into a lot of the issues that my plots in the exploration ran into where they would all be put in relative to one another and ended up looking really weird due to the y axis.

#Results

Overall, we reject the null hypothesis that there was no relationship between `Element`, `Type`, and the `Concentration_ppm`. As we found that both the `Element` (F = 26.44, p = 2.55 E-16)and `Type` (F = 32.99, p = 2.2e -16) had a significant and positive effect on the `Concentration_ppm`.

Again I'm not sure about this. I tried to model it based off the key from problem set 3 but the two-way anova doesn't match it very well.
